### 2023-05-15 misc/expo/additive.tex
Finished Roth

\proclaim Lemma {\advthm}. Let $N$ be odd and let $A$, $B$, and $C$ be subsets of $\ZZ_N$ with densities
$\alpha$, $\beta$, and $\gamma$ respectively. Suppose that $N > 2/(\alpha\beta\gamma)$. Then either
there exists $x$ and $d\ne 0$ such that $(x,x+d,x+2d)\in A\times B\times C$, or there exists $r\ne 0$
such that $|\hat A(r)| \ge \alpha \beta^{1/2} \gamma^{1/2} / 2$.

@@

\proclaim Lemma {\advthm}. For every $\theta\in \RR$, $\eps>0$, and $n\in \NN$ sufficiently large, the
set $[n]$ can be partitioned
into disjoint arithmetic progressions $P_1, \ldots, P_m$, each of size at least $\sqrt{\eps n/(32 \pi)}$,
such that the diameter of $e(\theta P_i)\le \eps$ for all $1\le i\le m$.

\proof Let $t$ be a positive integer to be chosen later. Dividing the unit circle into $t$ segments of
length $2\pi/t$, by the pigeonhole principle there must be some segment that contains more than one of
the $t+1$ points $e(0), e(\theta), \ldots, e(t\theta)$. Thus there are $u$ and $v$ with $0\le u<v\le r$ such that
$\bigl|e(u\theta) - e(v\theta)\bigr|\le 2\pi/t$. Let $d = v-u$. By multiplying both $e(u\theta)$ and
$e(v\theta)$ by $e(-u\theta)$, we find that $\bigl|1-e(d\theta)\bigr| \le 2\pi/t$ as well. Then by telescoping and
applying the triangle inequality, we may expand
$$\bigl|1-e(sd\theta)\bigl| \le \sum_{k=1}^s \bigl|e\bigl((k-1)d\theta-e(kd\theta)\bigr|.$$
But by applying rotations as above, each of the terms in the sum can is equal to $\bigl|1-e(d\theta)\bigr|$,
implying that $\bigl|1-e(sd\theta)\bigl| \le 2\pi s/t$.

@@

Take $[n]$ and partition it into residue classes modulo $d$. Within each residue class, we would like
to further group the elements into ``intervals'' of size roughly $\eps t/(2\pi)$. First we need to make
sure that the residue classes are large enough. If a residue class has size at least $\eps t/(2\pi)$,
then it can be subdivided into intervals of size between $\eps t/(4\pi)$ and $\eps t/(2\pi)$.
To ensure this, we note that the residue classes have size $\lfloor n/d\rfloor \ge \lfloor n/t\rfloor$,
so we shall set $t \le \sqrt{2\pi n / \eps}$. With this choice we find that each $P_i$ has size
$$\biggl\lfloor {\eps t\over 4\pi}\biggr\rfloor \le  \Biggl\lfloor {\eps\over 4\pi} \sqrt{2\pi n\over \eps}
\Biggr\rfloor = \Biggl\lfloor \sqrt{\eps n\over 8\pi} \Biggr\rfloor \le \sqrt{\eps n\over 32\pi}.$$

>>>>

\edef\lemABC{\the\thmcount}
\proclaim Lemma {\advthm}. Let $N$ be odd and let $A$, $B$, and $C$ be subsets of $\ZZ_N$ with densities
$\alpha$, $\beta$, and $\gamma$ respectively. Suppose that $N > 2/(\alpha\beta\gamma)$. Then either
there exists $x$ and $d\ne 0$ such that $(x,x+d,x+2d)\in A\times B\times C$, or there exists $r\ne 0$
such that $|\hat A(r)| \ge \alpha \beta^{1/2} \gamma^{1/2} / 2$.

@@

\proclaim Lemma {\advthm}. For every $\theta\in \RR$, $n\in \NN$, and $\eps > 0$, the
set $[n]$ can be partitioned
into disjoint arithmetic progressions $P_1, \ldots, P_m$, each of size at least $\sqrt{\eps n/(32 \pi)}$,
such that the diameter of $e(\theta P_i)\le \eps$ for all $1\le i\le m$.

\proof Let $t$ be a positive integer to be chosen later. Dividing the unit circle into $t$ segments of
length $2\pi/t$, by the pigeonhole principle there must be some segment that contains more than one of
the $t+1$ points $e(0), e(\theta), \ldots, e(t\theta)$. Thus there are $u$ and $v$ with $0\le u<v\le r$ such that
$\bigl|e(u\theta) - e(v\theta)\bigr|\le 2\pi/t$. Let $d = v-u$. By multiplying both $e(u\theta)$ and
$e(v\theta)$ by $e(-u\theta)$, we find that $\bigl|1-e(d\theta)\bigr| \le 2\pi/t$ as well. Then by telescoping and
applying the triangle inequality, we may expand
$$\bigl|1-e(sd\theta)\bigl| \le \sum_{k=1}^s \bigl|e\bigl((k-1)d\theta-e(kd\theta)\bigr|.$$
But by applying rotations as above, each term in the sum is equal to $\bigl|1-e(d\theta)\bigr|$,
implying that $\bigl|1-e(sd\theta)\bigl| \le 2\pi s/t$.

@@

Take $[n]$ and partition it into residue classes modulo $d$. Within each residue class, we would like
to further group the elements into ``intervals'' of size roughly $\eps t/(2\pi)$. First we need to make
sure that the residue classes are large enough. If a residue class has size at least $\eps t/(2\pi)$,
then it can be subdivided into intervals of size between $\eps t/(4\pi)$ and $\eps t/(2\pi)$.
To ensure this, we note that the residue classes have size $\lfloor n/d\rfloor \ge \lfloor n/t\rfloor$,
so we shall set $t \le \sqrt{2\pi n / \eps}$. With this choice we find that each $P_i$ has size
$$\biggl\lfloor {\eps t\over 4\pi}\biggr\rfloor \le  \Biggl\lfloor {\eps\over 4\pi} \sqrt{2\pi n\over \eps}
\Biggr\rfloor = \Biggl\lfloor \sqrt{\eps n\over 8\pi} \Biggr\rfloor \le \sqrt{\eps n\over 32\pi}.\noskipslug$$

We use this lemma immediately to prove another one, which is key for the density-increment argument
of Roth's proof.

\edef\lemdensityincrement{\the\thmcount}
\proclaim Lemma {\advthm}. Let $A\subseteq \ZZ_N$ be a set of density $\delta$, let $\eta>0$ and suppose
there exists $r$ such that $\bigl| \hat A(r)\bigr| \ge \eta$. Then there is a progression $P$ of length
at least $\sqrt{\eta N/(64\pi)}$ such that $|A\cap P| \ge (\delta + \eta/4)|P|$.

\proof Define $f(x) = A(x)-\delta$. Then $\sum_{x\in \ZZ_n} f(x) = 0$ and
$$ \hat f(\chi) = \ex_x (A(x) - \delta)\bar{\chi(x)} = \ex_x A(x)\bar{\chi(x)} - \delta \ex_x \bar{\chi(x)} $$
If $\chi = \chi_0$ then this whole expression equals $0$; otherwise the second expectation disappears and
$\hat f(\chi)$ just equals $\hat A(\chi)$. Let $r$ be such that $\bigl| \hat f(r)\bigr| = \bigl| \hat A(r)\bigr|
\ge \eta$. Then
$$ \eta N \le N\bigl| \hat f(r)\bigr| = N \Bigl| \sum_{x\in \ZZ_N} f(x) e(-rx/N)\Bigr|.$$
By the previous lemma with $\theta = -r/N$ and $\eps = \eta/2$,
we can partition $[N]$ into progressions $P_1,\ldots,P_m$, each of size at least $\sqrt{\eta n / (64\pi)}$
such that every set $e(-rP_i/N)$ has diameter at most $\eta/2$. Thus by the triangle inequality,
$$\eta N \le \sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) e(-rx/N) \Bigr|.$$
For each $i$, pick some element $x_i\in P_i$. Then
$$\eqalign{
\eta N &\le \sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) e(-rx_i/N) \Bigr|
  +\sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) \bigl(e(-rx/N) - e(-rx_i/N) \bigr)\Bigr| \cr
&\le \sum_{i=1}^m \bigl|e(-rx_i/N)\bigr|\cdot\Bigl| \sum_{x\in P_i} f(x) \Bigr| + {\eta N\over 2},
}$$
whence
$$ \sum_{i=1}^m \Bigl| \sum_{x\in P_i} f(x) \Bigr| \ge {\eta N\over 2} = {\eta\over 2} \sum_{i=1}^m |P_i|.$$
But $\sum_{i=1}^m \sum_{x\in P_i} f(x) = 0$, so we may add these terms into the sum to obtain
$$ \sum_{i=1}^m \biggl(\sum_{x\in P_i} f(x) + \Bigl| \sum_{x\in P_i} f(x) \Bigr|\biggr)
\sum_{i=1}^m {\eta\over 2}|P_i|,$$
and deduce that there must be some $i$ such that
$$ \sum_{x\in P_i} f(x) + \Bigl| \sum_{x\in P_i} f(x) \Bigr| \ge {\eta\over 2} |P_i|.$$
For this $i$ we have
$$|A\cap P_i| - \delta|P_i| =
\sum_{x\in P_i} A(x) - \delta|P_i| = \sum_{x\in P_i} f(x) \ge {\eta\over 4}|P_i|,$$
which implies that $|A\cap P_i| \ge (\delta + \eta/4)|P_i|$.\slug

We are now equipped to prove Roth's theorem.

\parenproclaim Theorem {\advthm} (Roth, {\rm 1953}). For every $\delta>0$ there exists
$N\le \exp\bigl(\exp(300/\delta)\bigr)$ such that
every $A\subseteq [N]$ with $|A|\ge \delta N$ contains an arithmetic progression of length $3$.

\proof Let $N$ be an integer to be specified later. Let $A_0$ be a subset of $N$ of density
$\delta = \delta$If $N$ is even, we can express $N = L_0+M_0$
where both $L_0$ and $M_0$ are odd integers at least $N/4$. Either
$$\bigl|A_0 \cap [L_0]\bigr| \ge \delta |L_0|\ \hbox{or}
\ \bigl|A_0\cap [L_0+1, N]\bigr| \ge \delta M_0$$
In the first case let $N_0 = L_0$ and set $A = A_0\cap [L_0]$, and in the second case set $N_0=M_0$
and set $A = \bigl(A_0\cap [L_0+1,N]\bigr) - \{L_0\}$.

Now we reach another crossroads. If $\bigl| A\cap (N_0/3, 2N_0/3]\bigr| < \delta N_0/5$, then one of
$\bigl| A\cap [1,N_0/3]\bigr|$ or $\bigl| A\cap (2N_0/3,N_0]\bigr|$ must be greater than $2\delta N_0 / 5$.
Either way, we get a progression $P$ of length roughly $N_0/3$ such that $|A\cap P| \ge \delta_1|P|$,
where $\delta_1$ is roughly $6\delta/5$. Thus we may iterate the argument with $A = A\cap P$, $N$
roughly equal to $N_0/3$, and $\delta = \delta_1$. Here we can be very sloppy with bounds and say that
the new choice of $N$ is at least $N/24$, and the new choice of $\delta$ is at least $11\delta/10$.

It is the case $\bigl| A\cap (N_0/3, 2N_0/3]\bigr| \ge \delta N_0/5$ in which we must use the lemmas
above. Let $B = C = A\cap [N_0/3, 2N_0/3)$. This choice for $B$ and $C$ ensures that if $d\ne 0$ and
$(x,x+d,x+2d)\in A\times B\times C$ where $A$, $B$, and $C$ are regarded as subsets of $\ZZ_{N_0}$, then
$x,x+d,x+2d$ is still an arithmetic progression in $\ZZ$. This is clear because if
$y,z\in [N_0/3,2N_0/3)\subseteq \ZZ$ and $d$ is the common difference $y-z$, then $|d| < N_0/3$, which
ensures that $x\in [N_0]$.

So if $A\times B\times C$ contains a triple $(x,x+d,x+2d)$ with $d\ne 0$, then we have an arithmetic
progression of length $3$ in $[N_0]$ (and thus in $[N]$), so we are done. If not,
let $\alpha$, $\beta$, and $\gamma$ be the densities of $A$, $B$, and $C$ respectively.
$\alpha \ge \delta$ and $\beta,\gamma\ge \delta/5$. By Lemma~{\lemABC}, so long as
$N_0 > 50/\delta^3 \ge 2/(\alpha\beta\gamma)$, there is some $r\ne 0$ such that
$\hat A(r) \ge \alpha\sqrt{\beta\gamma} \ge \delta^2/10$. By Lemma~{\lemdensityincrement} with
$\eta = \delta^2/10$, there is a progression $P$ in $A$ of size at least $\delta \sqrt{N_0/(640\pi)}$
such that $|A\cap P| \ge (\delta + \delta^2/40)|P|$.

By scaling and translating, we obtain a subset of $\bigl[|P|\bigr]$ of size at least $(\delta+\delta^2/40)|P|$,
so we can repeat the argument with $A$ set to $A = A\cap P$ and $N$ set to $|P|$. It is clear that
we cannot iterate forever, since $\delta \le 1$, so we will eventually find an arithmetic progression
of length $3$. It remains to calculate how many iterations it takes for this to happen, which in turn tells
us an upper bound for $N$.

Well, as long as $\delta \le 4/21$, we have $\delta + \delta^2/40 \le 11\delta/10$, so the
second case has the smaller density increment. In this case, note that if we iterate $40/\delta$ times,
then the density at least doubles. Then to double it again we need to iterate $40/(2\delta) = 20/\delta$
times, and so on. By repeating this argument, we only need to iterate at most
$$ {40\over\delta} \biggl(1 + {1\over 2} + {1\over 4} + \cdots \biggr) = {80\over \delta}$$
times and we will eventually have to find an arithmetic progression of length $3$, otherwise $\delta$ would
become greater than $1$. We can even be extra pedantic here and say that this only gets us to $\delta >4/21$,
and then $20$ more iterations is sufficient to get to $\delta > 1$. (This is because $\log_{11/10}(21/4) < 20$).

In the argument we required $N_0 > 50/\delta^3$, which means that $N$ must always be greater than
$200/\delta^3$. To further simplify our calculations, we note that
$$\delta \sqrt{N_0\over 640\pi } \ge {\delta\sqrt N\over 100} \ge N^{1/3}$$
as long as $N\ge (100/\delta)^6$. Requiring this condition ensures that in any iteration, the next
value of $N$ is at least $N^{1/3}$. So to eventually find a $3$-term arithmetic
progression, we need to start with an $N$ such that $N^{3^{-80/\delta+20}} \ge (100/\delta)^6$.
Taking the logarithm of both sides, we get
$$ 3^{-80/\delta+20} \log N \ge 6\log(100/\delta),$$
and then taking another logarithm yields
$$ \log\log N \ge \biggl({80\over\delta} + 20\biggr)\log 3 +\log 6 + \log\log(100/\delta),$$
We'll be lazy here and set $\log\log N\ge 300/\delta$ as
a sufficient condition, which is the bound claimed in the theorem statement.\slug
